\section{Installation}

\subsection{Windows}

The use the GROOVE tool set under windows, download the bin+lib package from the download site explained above and unzip it to any location on your computer. We will refer to the destination folder as GROOVE\_PATH. Under GROOVE\_PATH, you will find a directory \emph{bin} which contains some batchfiles, that refer to the tools contained by the GROOVE tool set (Editor.bat, Simulator.bat, Generator.bat and Imager.bat). Inside the batchfiles, a variable GROOVE\_BIN\_DIR is set. Make sure the value it is set to is your local GROOVE\_PATH.

Then make sure you have a JRE (Java Runtime Environment) installed and that you have it in your PATH. You can download a JRE from \url{http://java.sun.com}.

Now double-click on a batch-file to run the desired tool.

\subsection{NIX}

The use the GROOVE tool set under windows, download the bin+lib package from the download site explained above and unzip it to any location on your computer. We will refer to the destination folder as GROOVE\_PATH. Under GROOVE\_PATH, you will find a directory \emph{bin} which contains some executable shell-scripts, that refer to the tools contained by the GROOVE tool set (Editor, Simulator, Generator and Imager). Inside the shell-scripts, a variable GROOVE\_BIN\_DIR is set. Make sure the value it is set to is your local GROOVE\_PATH.

Then make sure you have a JRE (Java Runtime Environment) installed and that you have it in your PATH. You can download a JRE from \url{http://java.sun.com}.

Now execute a shell-script to run the desired tool.
